---
title: "Biolog Cleaning Script"
author: "Adrian Noecker"
date: "September 2023"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

COLLABORATORS: Moriah Young, Phoebe Zarnetske, Mark Hammond, Taylor Ulbrich
DATA INPUT:  csv from the shared Google drive  
DATA OUTPUT: Code and Rmd are in the scripts folder in Github  
PROJECT: REX

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading in data
```{r}
# Clear all existing data
rm(list=ls())

#load packages
library(car)
library(tidyr)
library(dplyr)
library(vegan)
library(tibble)
library(plyr)
library(readr)
library(tidyverse)

# Set working directory from .Renviron
dir <- Sys.getenv("DATA_DIR")
list.files(dir)

## Read in data
#time 0
time0 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/time0/")
time0 = list.files(path=time0, pattern="*.csv", full.names=TRUE)
time0_csv = ldply(time0, read_csv)

#day 1
day1 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day1/")
day1 = list.files(path=day1, pattern="*.csv", full.names=TRUE)
day1_csv = ldply(day1, read_csv)

#day 2
day2 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day2/")
day2 = list.files(path=day2, pattern="*.csv", full.names=TRUE)
day2_csv = ldply(day2, read_csv)

#day 3
day3 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day3/")
day3 = list.files(path=day3, pattern="*.csv", full.names=TRUE)
day3_csv = ldply(day3, read_csv)

#day 4
day4 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day4/")
day4 = list.files(path=day4, pattern="*.csv", full.names=TRUE)
day4_csv = ldply(day4, read_csv)

#day 5
day5 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day5/")
day5 = list.files(path=day5, pattern="*.csv", full.names=TRUE)
day5_csv = ldply(day5, read_csv)

#day 6
day6 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day6/")
day6 = list.files(path=day6, pattern="*.csv", full.names=TRUE)
day6_csv = ldply(day6, read_csv)

#day 7
day7 = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/day7/")
day7 = list.files(path=day7, pattern="*.csv", full.names=TRUE)
day7_csv = ldply(day7, read_csv)

#meta data
meta = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/")
meta = list.files(path=meta, pattern="*.csv", full.names=TRUE)
meta_csv = ldply(meta, read_csv)

#meta data
hour = file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L0/")
hour = list.files(path=hour, pattern="*.csv", full.names=TRUE)
hour_csv = ldply(hour, read_csv)
```

#Merge
```{r}
#changing plate ID to plate in raw reads for merging
time0mod <- time0_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day1mod <- day1_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day2mod <- day2_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day3mod <- day3_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day4mod <- day4_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day5mod <- day5_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day6mod <- day6_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))
day7mod <- day7_csv %>% separate('Plate ID', c('Intials', 'Plate_word', 'Plate', 'day'))

#Making plate columns numeric
time0mod$Plate = as.numeric(as.character(time0mod$Plate))
day1mod$Plate = as.numeric(as.character(day1mod$Plate))
day2mod$Plate = as.numeric(as.character(day2mod$Plate))
day3mod$Plate = as.numeric(as.character(day3mod$Plate))
day4mod$Plate = as.numeric(as.character(day4mod$Plate))
day5mod$Plate = as.numeric(as.character(day5mod$Plate))
day6mod$Plate = as.numeric(as.character(day6mod$Plate))
day7mod$Plate = as.numeric(as.character(day7mod$Plate))

#merging each day with meta
time0merge<-merge(meta_csv, time0mod, by=c("Plate","Well"))
day1merge<-merge(meta_csv, day1mod, by=c("Plate","Well"))
day2merge<-merge(meta_csv, day2mod, by=c("Plate","Well"))
day3merge<-merge(meta_csv, day3mod, by=c("Plate","Well"))
day4merge<-merge(meta_csv, day4mod, by=c("Plate","Well"))
day5merge<-merge(meta_csv, day5mod, by=c("Plate","Well"))
day6merge<-merge(meta_csv, day6mod, by=c("Plate","Well"))
day7merge<-merge(meta_csv, day7mod, by=c("Plate","Well"))

#merging all readings to one
plates <- rbind(time0merge, day1merge, day2merge, day3merge, day4merge, day5merge, day6merge, day7merge)


#change date to hour 
ecoplates <- merge(x=plates,y=hour_csv, by.x=c('Reading Date/Time'), 
      by.y=c('Reading Date/Time'))


#confirm that we have the same number of rows 
length(rownames(day7_csv)) = 960
length(rownames(meta_csv)) =960
length(rownames(plates)) = 7680
length(rownames(ecoplates)) = 7680

```

#Cleaning
```{r}
# remove columns because we don't need them for analysis
clean_ecoplates <- ecoplates %>%
  select(-Intials, -Plate_word, -Biolog_Plate_Wells_Location)


# Convert any data in the Delta column that is <0 to 0 
clean_ecoplates$Delta[clean_ecoplates$Delta < 0]<-0
```

# standardize each substrate absorance value 
## First, standardize by substract the time = 0 value from every other time point for each individual well
```{r}
# Filter for just the Hour0
hour0 <- filter(clean_ecoplates, Hour == 0)
dim(hour0)

# Keep just the Hour0 Delta, Plate and Well Columns and change Delta column name
hour0mod <- select(hour0, Plate, Well, Delta)
colnames(hour0mod)[3] <- "Delta_Hr0" ##Delta to Delta_Hr0 

# Merge biolog dataset with all times and biolog_Hr0 which has the separate column for hour0 
biolog <- merge(clean_ecoplates, hour0mod, by=c("Plate", "Well"))

# Now substract the PlateWell absorbance at Hr0 from every other Hr
biolog$Delta_SubHr0<- biolog$Delta - biolog$Delta_Hr0

# in some cases, the absorbance at time = 0 was greater than values at other hours, leading to negative difference when we normalized. 
# SO, make everything that is <0 == 0 now after normalization 
# Convert any data in the Delta_SubHr0 column that is <0 to 0 
biolog$Delta_SubHr0[biolog$Delta_SubHr0 < 0] <- 0

```

## Second, Standardize by subtracting the water absorbance value for each sample at every time point
```{r}
#Substracting water using same method as Hour 0
water <- filter(clean_ecoplates, Well_C_Source == 'Water')

#removing unneeded columns
watermod <- select(water, Plate, Sample_ID, Delta, Hour)
colnames(watermod)[3] <- "Delta_Water" ##Delta to Delta_Water 

# Merge biolog dataset with all times and samples which has the separate column for water 
biolog1 <- merge(biolog, watermod, by=c("Plate", "Sample_ID", "Hour"))

# Now substract the PlateWell absorbance at water from every other sample at each time
biolog1$Delta_MinusWater<- biolog1$Delta - biolog1$Delta_Water

# Again, Convert any data in the Delta_Blank column that is <0 to 0 
biolog1$Delta_MinusWater[biolog1$Delta_MinusWater < 0] <- 0
```

# Reformat dataframe into and abundance matrix and metadata for diversity assessments
```{r}
# first remove the unnecessary columns 
biolog2 = select(biolog1, Delta_MinusWater, Delta, Delta_SubHr0, Well_C_Source, Hour, Plot_ID, Sample_ID, Plate, Well, Biolog_column_ID) # remove all other columns 
View(biolog2)

# save this is a completed, standardized data 
write.csv(biolog2, file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L1/Biolog_data_L1.csv"), row.names = FALSE)

# now reshape this and spread the data into wide format
# spread(df, variable-to-be-columns, variable-to-fill-matrix(absorbance), fill = 0(this fills in 0s for any missing data))
# http://environmentalcomputing.net/reshaping-data/
biolog3 = select(biolog2, Delta_MinusWater, Well_C_Source, Hour, Plot_ID, Sample_ID, Biolog_column_ID) # remove all other columns 
data_reshape <- spread(biolog3, Well_C_Source, Delta_MinusWater)
View(data_reshape)

# Create an abundance table from this newly reshaped dataframe
# remove water because this was only a blank/control and shouldn't be included in diversity indices 
Absorb <- select(data_reshape, -Water)
View(Absorb)

write.csv(Absorb, file.path(dir, "/microbes/Biolog EcoPlates/T7_warmx/L1/Biolog_data_NOWater_L1.csv"), row.names = FALSE)
```

# reformatting and checking absorbances
# Note: the "biolog" dataframe is in wide format, and is the format that the data is output from the plate reader. The "standard" data frame is in narrow format.
```{r}
#removing water from standard biolog
standardnowater <- standardbiolog %>% filter(!Well_C_Source %in% c("Water"))

#changing row name and reformatting, combine hour and ID to one column
biologdf <- biolognowater %>%
  unite(SampleID_Hour, Biolog_column_ID, Hour, sep ="_", remove = FALSE)

standarddf <- standardnowater %>%
  unite(SampleID_Hour, Biolog_column_ID, Hour, sep ="_", remove = FALSE)

#renaming new column
biologdf$rowname <- biologdf$SampleID_Hour

# Note from Moriah: is there a reason why the first column is converted to the row names??
#changing column to row
biologmatrix <- column_to_rownames(biologdf, "rowname")

#Check control absorbances (all should == 0 if no contamination)
biologcontrol <- biologmatrix %>% filter(Sample_ID=="BIOLOG CONTROL")
####Tween 40, Tween 80, D Xylose, X4 Hydroxy Benzoic Acid, and X2 Hydroxy Benzoic Acid larger than 0.01 in a few controls, most hour 0 in plates 5, 7, 9, and 10 so maybe just the initial color of well and not contamination??

standardcontrol <- standarddf %>% filter(Sample_ID=="BIOLOG CONTROL")

#exclude controls for analysis
biologsamples <- biologmatrix %>% filter(!Sample_ID %in% c("BIOLOG CONTROL"))

standardsamples <- standarddf %>% filter(!Sample_ID %in% c("BIOLOG CONTROL"))

#import meta data to get treatment type
meta <- read.csv(file.path(dir, "/REX_warmx_metadata.csv"))

#get rid of unneeded columns
meta1 <- meta %>% dplyr::select(Subplot_Descriptions, Unique_ID, Rep, Footprint_Location)

# Note from Moriah: when we merge these two together, we lose the matrix formatting - may need to run "column_to_rownames" function again on new dataframe to make SampleID_Hour row names
#combine meta and sample df
biologsamples1 <- merge(x=biologsamples, y=meta1, by.x=c('Plot_ID'), 
      by.y=c('Unique_ID'))

standardsamples1 <- merge(x=standardsamples, y=meta1, by.x=c('Plot_ID'), 
      by.y=c('Unique_ID'))

#change column order
biologsamples2 <- biologsamples1 %>% relocate(Subplot_Descriptions, .after = Biolog_column_ID) %>% 
        relocate(Rep, .after = Plot_ID) %>% 
        relocate(Footprint_Location, .after = Rep)

# upload new data frame to L2 folder in warmx 
write.csv(biologsamples2, file.path(dir, "/microbes/Biolog EcoPlates (1)/T7_warmx/L2/biolog_clean_L1.csv"), row.names = FALSE)

```


